---
title: "Chapter 3"
output:
  pdf_document:
    number_sections: true
    toc: true
    extra_dependencies: ["amsthm", "amsmath"]
  html_notebook: default
---


```{r}
library(sf)
library(terra)
library(dplyr)
library(spData)
data(us_states)
data(us_states_df)
```

E1. Create a new object called `us_states_name` that contains only the `NAME` column from the `us_states` object using either base R (`[`) or tidyverse (`select()`) syntax.
What is the class of the new object and what makes it geographic?

```{r}
us_states_name <- us_states[, "NAME"]
class(us_states_name)
```

E2. Select columns from the `us_states` object which contain population data.
Obtain the same result using a different command (bonus: try to find three ways of obtaining the same result).
Hint: try to use helper functions, such as `contains` or `matches` from **dplyr** (see `?contains`).

```{r}
# E2
us_states_pop1 <- us_states[, c("total_pop_10", "total_pop_15")]

us_states_pop1
```

```{r}
us_states |> dplyr::select(total_pop_10, total_pop_15)
```

E3. Find all states with the following characteristics (bonus find *and* plot them):

- Belong to the Midwest region.

```{r}
# E3
us_states_midwest <- us_states[us_states$REGION=="Midwest",]
plot(st_geometry(us_states_midwest))
```

```{r}
#E3
plot(st_geometry(us_states |> 
  filter(REGION == "Midwest")))
```

- Belong to the West region, have an area below 250,000 km^2^ *and* in 2015 a population greater than 5,000,000 residents (hint: you may need to use the function `units::set_units()` or `as.numeric()`).

```{r}
# E3
us_states_west <- us_states[us_states$REGION=="West",]
us_states_west <- us_states_west[us_states_west$AREA < units::set_units(250000, "km^2"), ]
us_states_west <- us_states_west[us_states_west$total_pop_15 > 5000000, ]
us_states_west
plot(st_geometry(us_states_west))
```

- Belong to the South region, had an area larger than 150,000 km^2^ or a total population in 2015 larger than 7,000,000 residents.

```{r}
us_states_south <- us_states |> filter(REGION=="South", AREA > units::set_units(150000, "km^2") | total_pop_15 > 7000000)
plot(st_geometry(us_states_south))
```

E4. What was the total population in 2015 in the `us_states` dataset?
What was the minimum and maximum total population in 2015?

```{r}
#E4
us_states |> summarize(total_pop = sum(total_pop_15),
                       min_pop = min(total_pop_15),
                       max_pop = max(total_pop_15))
```

E5. How many states are there in each region?

```{r}
#E5
# n() gives the group size 
us_states |> 
  group_by(REGION) |>
  summarise(nr_of_states = n())
```

E6. What was the minimum and maximum total population in 2015 in each region?
What was the total population in 2015 in each region?

```{r}
#E6
summarize(group_by(us_states, REGION), 
          min_pop=min(total_pop_15),
          max_pop=max(total_pop_15),
          tot_pop=sum(total_pop_15))
```

E7. Add variables from `us_states_df` to `us_states`, and create a new object called `us_states_stats`.
What function did you use and why?
Which variable is the key in both datasets?
What is the class of the new object?

```{r}
#E7
us_states_stats <- us_states |>
  left_join(us_states_df, by = c("NAME"="state"))
class(us_states_stats)
```

E8. `us_states_df` has two more rows than `us_states`.
How can you find them? (hint: try to use the `dplyr::anti_join()` function)

```{r}
#E8
dplyr::anti_join(us_states_df, st_drop_geometry(us_states),
                 by = c("state" = "NAME"))
```

E9. What was the population density in 2015 in each state?
What was the population density in 2010 in each state?

```{r}
#E9
us_pop_density_15 <- us_states$total_pop_15 / us_states$AREA
us_pop_density_15
```

or

```{r}
#E9
us_states2 <- us_states |>
  mutate(pop_dens_15 = total_pop_15/AREA,
         pop_dens_10 = total_pop_10/AREA)

us_states2$pop_dens_15
```

E10. How much has population density changed between 2010 and 2015 in each state?
Calculate the change in percentages and map them.

```{r}
#E10
us_states2$pop_dens_change <- 
  us_states2$pop_dens_15 / us_states2$pop_dens_10

plot(us_states2[, "pop_dens_change"] )
```

E11. Change the columns' names in `us_states` to lowercase. (Hint: helper functions - `tolower()` and `colnames()` may help.)

```{r}
us_states |> 
  setNames(tolower(colnames(us_states)))
```

or

```{r 03-ex-e11}
us_states %>%
  setNames(tolower(colnames(.)))
?setNames
```

E12. Using `us_states` and `us_states_df` create a new object called `us_states_sel`.
The new object should have only two variables - `median_income_15` and `geometry`.
Change the name of the `median_income_15` column to `Income`.

```{r}
us_states_df
```


```{r}
us_states_sel <- left_join(us_states, us_states_df, by = c("NAME" = "state"))
us_states_sel <- dplyr::select(us_states_sel, Income=median_income_15)
us_states_sel
```

or

```{r}
us_states_sel <- us_states |>
  left_join(us_states_df, by = c("NAME" = "state")) |>
  dplyr::select(Income = median_income_15)
```

E13. Calculate the change in the number of residents living below the poverty level between 2010 and 2015 for each state. (Hint: See ?us_states_df for documentation on the poverty level columns.)
Bonus: Calculate the change in the *percentage* of residents living below the poverty level in each state.

```{r}

```







